---
title: "CDS 101 – Final Project Report"
author: "Group Name / Members"
date: "`r format(Sys.Date(), '%B %d, %Y')`"
output:
  html_document:
    css: "gmu-cds101.css"
    toc: true
    toc_depth: 3
    number_sections: true
  pdf_document:
    toc: true
---

> Replace this template text with your own writing, keeping the headings and overall structure aligned with the **CDS 101 project rubric**.

# 1. Problem Definition

- The main problem or question we are investigating is if time of day impacts the sale of hot chocolate/cocoa drinks?.  
- This question is interesting because the general recommendation is to avoid caffeine later in the day.  Knowing this, we would expect to see hot cocoa sales improve later in the day as an uncaffeinated option.  
- Methods: Two-way hypothesis testing because we chose to examine categorical variables and want to note ANY change, positive or negative. 

Null hypothesis: Time of day does not impact the sales of hot chocolate/cocoa drinks. 

Alternative hypothesis: Time of day changes hot chocolate/cocoa drink sales. 

Expected result: We expect hot chocolate/cocoa sales to take up a larger portion of sales later in the day. 

# 2. Data Acquisition & Description

- The name of the dataset is "Coffee shop sales" and it was retrieved from Kaggle. 
- The file was downloaded directly from Kaggle.  
- 1. transaction_id

Unique identifier for each sales transaction.
Type: Integer
Units: None

2. transaction_date

The calendar date when the transaction occurred.
Type: Date
Units: YYYY-MM-DD

3. transaction_time

The time of day when the transaction was recorded.
Type: Time (POSIX time)
Units: HH:MM:SS

4. transaction_qty

Number of units/items purchased in the transaction.
Type: Integer
Units: Count

5. store_id

Unique code identifying the store where the transaction took place.
Type: Integer
Units: None

6. store_location

The name or geographic location of the store.
Type: Character (categorical)
Units: None

7. product_id

Unique identifier for the product sold.
Type: Integer
Units: None

8. unit_price

Price of a single unit of the product sold.
Type: Numeric
Units: Currency (USD)

9. product_category

Broad category of the product (e.g., Coffee, Tea, Hot cocoa).
Type: Character (categorical)
Units: None

10. product_type

More specific product grouping within each category (e.g., Latte, Iced Coffee).
Type: Character (categorical)
Units: None

11. product_detail

Detailed description of the product (e.g., “Vanilla Latte – Medium”).
Type: Character (categorical / descriptive text)
Units: None  

- The file contains 149116 rows and 11 columns.  
- There shouldn't be any ethical, privacy, or bias considerations.

```{r setup-data, echo=FALSE, message=FALSE, warning=FALSE}
library(readxl)

data <- read_excel("Coffee Shop Sales.xlsx")
```

```{r}
dim(data)
names(data)
```

```{r}
head(data, 10)
```

# 3. Data Cleaning & Preprocessing

Describe the steps you took to make the data usable, for example:

- Handling missing values (drop, impute, etc.).  
- Fixing types (dates, numeric vs. categorical).  
- Removing outliers (if done).  
- Creating new features (feature engineering).  
- Any scaling or encoding.

```{r cleaning, eval=FALSE}
library(dplyr)

coffee_sales <- data %>%
  select(transaction_time,
         transaction_qty,
         product_category,
         product_type,
         product_detail)
```

# 4. Exploratory Data Analysis (EDA)

This section maps directly to the **EDA** part of the rubric:

- Summary statistics  
- Visualizations  
- Interpretation connected to the research question  

```{r eda-summary, message=FALSE, warning=FALSE}
# library(dplyr)
# summary(data_clean)
```

```{r eda-plot-example, message=FALSE, warning=FALSE, fig.cap="Example histogram of a numeric variable (EDA rubric)."}
# library(ggplot2)
# ggplot(data_clean, aes(x = some_numeric_variable)) +
#   geom_histogram(binwidth = 5, color = "white") +
#   labs(x = "Some Variable", y = "Count", title = "Distribution of Some Variable")
```

# 5. Visualization Quality and Storytelling

Use this section to satisfy the **Visualization Quality** rubric criterion:

- Explain why your plot types are appropriate.  
- Comment on labels, legends, colors, and overall readability.  
- Mention any steps you took to make plots accessible and interpretable.

# 6. Modeling Approach

Explain how you framed the problem and which models you chose:

- Type of task (regression, classification, etc.).  
- Baseline model or heuristic, if used.  
- Main model(s) chosen and why they are appropriate.

# 7. Model Implementation & Evaluation

This corresponds to the **Model Implementation & Evaluation** rubric criterion.

Describe:

- Features used.  
- Data splitting strategy (train/test or cross-validation).  
- Metrics used (accuracy, F1, RMSE, etc.).  
- Tables/plots summarizing performance.  
- A short interpretation of each metric/plot.

```{r modeling, eval=FALSE}
# Example structure:
# set.seed(123)
# train_index <- sample(seq_len(nrow(data_clean)), size = 0.8 * nrow(data_clean))
# train <- data_clean[train_index, ]
# test  <- data_clean[-train_index, ]
#
# model <- glm(target ~ ., data = train, family = binomial)
# preds <- predict(model, newdata = test, type = "response")
# # compute metrics...
```

# 8. Conclusions & Recommendations

Summarize the key takeaways:

- Answer your original research question(s) directly.  
- Highlight the most important patterns or relationships you found.  
- Discuss limitations (data size, bias, missing variables, etc.).  
- Suggest possible extensions or future work.

# 9. Code Quality & Reproducibility

Briefly document how someone else can reproduce your results:

- Which R scripts or Rmd files to run.  
- Any required R packages.  

```{r session-info, echo=FALSE}
sessionInfo()
```

# 10. References

List any references you used, such as:

- Dataset documentation.  
- Research papers or articles.  
- Tutorials or blog posts.

# Appendix (Optional)

Include extra plots, diagnostic checks, or model comparisons if needed.
